---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mrmsR

<!-- badges: start -->
[![R-CMD-check](https://github.com/searsmg/mrmsR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/searsmg/mrmsR/actions/workflows/R-CMD-check.yaml)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.22908308.svg)](https://doi.org/10.5281/zenodo.22908308)
<!-- badges: end -->

Tools for processing NOAA MRMS (Multi-Radar Multi-Sensor) gridded
precipitation data and generating watershed-scale rainfall statistics.

## Overview

`mrmsR` supports a five-step workflow for turning MRMS gridded precipitation
data into watershed-scale rainfall statistics:

1. **`downloadMRMS()`** — download MRMS GRIB2 files for a time range from the
   Iowa State Mesonet archive (or the NOAA AWS bucket, for `SHSRHeight`).
2. **`prepMRMS()`** — project and crop the raw GRIB2 rasters to a boundary,
   writing GeoTIFFs.
3. **`zonalMRMS()`** — compute the zonal mean precipitation over each
   catchment in the boundary for every raster, writing one CSV per timestep.
4. **`combineCSV()`** — combine those per-timestep CSVs into a single,
   datetime-sorted data frame.
5. **`accumulateMRMS()`** — convert the values to rainfall depths and total
   them by hour, day, or the whole record.

The package is designed for high-volume hydroclimatic workflows and supports
efficient batch and parallel processing.

## Installation

`mrmsR` is only available on GitHub. Install the latest version using
[pak](https://pak.r-lib.org/):

``` r
# install.packages("pak")
pak::pak("searsmg/mrmsR")
```

To install a specific release (for example, to match a version you've cited):

``` r
pak::pak("searsmg/mrmsR@v0.1.0")
```

## Example

The example below runs the full pipeline. Downloading requires a network
connection and can produce a lot of files, so that step is shown but not
run here. The remaining steps run on a small example dataset bundled with
the package (also used in `vignette("mrmsR")`), so every output below is
reproducible.

```{r setup}
library(mrmsR)
```

### Step 1: Downloading MRMS data

```{r download, eval = FALSE}
downloadMRMS(
  start = lubridate::ymd_hm("2025-01-01 02:00"),
  end = lubridate::ymd_hm("2025-01-01 06:00"),
  destination = "mrms_rasters",
  product = "RadarOnlyQPE"
)
```

`product` controls both the data source and the temporal resolution of the
files that get downloaded:

| Product              | Resolution | Source                    |
|-----------------------|------------|----------------------------|
| `RQI`                | 2 minutes  | Iowa State Mesonet archive |
| `MultiSensorQPE`     | hourly     | Iowa State Mesonet archive |
| `RadarOnlyQPE`       | hourly     | Iowa State Mesonet archive |
| `SurfacePrecipRate`  | 2 minutes  | Iowa State Mesonet archive |
| `SHSRHeight`         | 2 minutes  | NOAA MRMS AWS bucket       |

MRMS file names are in UTC. `start` and `end` can be in any time zone
(e.g., `lubridate::ymd_hm("2025-07-03 14:00", tz = "America/Chicago")`); they're
converted to UTC before files are requested. `destination` is created if it
doesn't exist.

`downloadMRMS()` returns a list of the `POSIXct` timestamps (in UTC) it could
not download. An empty list means every file in the range downloaded and
unzipped successfully.

### Example data

```{r example-data}
grib_dir <- system.file("extdata", "grib2", package = "mrmsR")
boundary <- system.file("extdata", "boundary", "catchments_all_lidar.shp", package = "mrmsR")

list.files(grib_dir)
```

### Step 2: Cropping rasters to a boundary

```{r prepMRMS, warning = FALSE}
tif_dir <- file.path(tempdir(), "mrmsR-readme-tif")
dir.create(tif_dir, showWarnings = FALSE)

prepMRMS(
  dir = grib_dir,
  output_dir = tif_dir,
  num_cores = 1,
  boundary = boundary
)

list.files(tif_dir)
```

```{r plot-raster}
r <- terra::rast(list.files(tif_dir, full.names = TRUE)[1])
terra::plot(r, main = names(r))
```

### Step 3: Zonal statistics per catchment

`zonalMRMS()` computes the mean precipitation within each catchment polygon
and writes one CSV per raster, named `extract_<year>_<doy>_<hour>_<min>.csv`.
The `catchment` column comes from the boundary's `site` attribute if it has
one; use `id_col` to pick a different attribute. Set `n_workers` above 1 to
process rasters in parallel.

```{r zonalMRMS}
csv_dir <- file.path(tempdir(), "mrmsR-readme-csv")
dir.create(csv_dir, showWarnings = FALSE)

zonalMRMS(
  raster_dir = tif_dir,
  output_dir = csv_dir,
  boundary = boundary,
  n_workers = 1
)

list.files(csv_dir)
```

### Step 4: Combining timesteps into one time series

```{r combineCSV}
combined <- combineCSV(csv_dir)

combined
```

`combined` is a normal data frame, so you can also work with it directly
using your usual `dplyr`/`tidyr` tools, e.g. to plot hyetographs.

### Step 5: Accumulating rainfall totals

`accumulateMRMS()` turns per-timestep values into rainfall depths (mm) and
totals them for each catchment by `"hour"`, `"day"`, or `"total"`. It
handles the unit conversion for each product: `SurfacePrecipRate` is a rate
in mm/hr every 2 minutes, while the hourly QPE products are already depths.
Use `tz` to total by local days instead of UTC days.

```{r accumulateMRMS}
accumulateMRMS(combined, product = "RadarOnlyQPE", by = "total")
```

`n_steps` counts the timesteps that went into each total and `expected_steps`
is how many a complete period has, so you can spot gaps in the data.

See `vignette("mrmsR")` for more detail on each step.

## Citation

If you use `mrmsR`, please cite it using its Zenodo DOI:

> Sears, M. mrmsR: Download and process MRMS data. Zenodo.
> <https://doi.org/10.5281/zenodo.22908308>

That DOI always points to the latest release. Each release also has its own
DOI, listed on the [Zenodo record](https://doi.org/10.5281/zenodo.22908308),
for citing the exact version you used. GitHub's **Cite this repository**
button gives the same citation in APA and BibTeX formats.

## Project status

`mrmsR` is in early development. Version 0.1.0 is the first release.

- Function names, arguments, and outputs may still change between versions
- Breaking changes are listed in [NEWS.md](NEWS.md)

### Planned improvements

- Performance optimization for large datasets
- Option to pull MRMS either by catchment (boundary) or point
- Review by others
