---
title: "Getting started with mrmsR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with mrmsR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mrmsR)
```

## Overview

`mrmsR` supports a four-step workflow for turning NOAA MRMS (Multi-Radar
Multi-Sensor) gridded precipitation data into watershed-scale rainfall
statistics:

1. **`downloadMRMS()`** — download MRMS GRIB2 files for a time range from the
   Iowa State Mesonet archive (or the NOAA AWS bucket, for `SHSRHeight`).
2. **`prepMRMS()`** — project and crop the raw GRIB2 rasters to a boundary,
   writing GeoTIFFs.
3. **`zonalMRMS()`** — compute the zonal mean precipitation over each
   catchment in the boundary for every raster, writing one CSV per timestep.
4. **`combineCSV()`** — combine those per-timestep CSVs into a single,
   datetime-sorted data frame.

This vignette walks through the whole pipeline. Downloading requires a
network connection and can produce a lot of files, so that step is shown but
not run here. The remaining steps run on a small example dataset bundled
with the package, so you can follow along and reproduce every output below.

## Step 1: Downloading MRMS data

```{r download, eval = FALSE}
downloadMRMS(
  start = lubridate::ymd_hm("2025-01-01 02:00"),
  end = lubridate::ymd_hm("2025-01-01 06:00"),
  destination = "mrms_rasters",
  product = "RadarOnlyQPE"
)
```

`product` controls both the data source and the temporal resolution of the
files that get downloaded:

| Product              | Resolution | Source                    |
|----------------------|------------|----------------------------|
| `RQI`                | 2 minutes  | Iowa State Mesonet archive |
| `MultiSensorQPE`     | hourly     | Iowa State Mesonet archive |
| `RadarOnlyQPE`       | hourly     | Iowa State Mesonet archive |
| `SurfacePrecipRate`  | 2 minutes  | Iowa State Mesonet archive |
| `SHSRHeight`         | 2 minutes  | NOAA MRMS AWS bucket       |

`downloadMRMS()` returns a list of the `POSIXct` timestamps it could not
download (for example if the archive doesn't have data for a given
timestep). An empty list means every file in the range downloaded and
unzipped successfully.

## Example data

The rest of this vignette uses a small example dataset installed with the
package: two hourly `RadarOnlyQPE` GRIB2 files (2025-01-01 02:00 and 03:00
UTC) and a shapefile of catchment boundaries.

```{r example-data}
grib_dir <- system.file("extdata", "grib2", package = "mrmsR")
boundary <- system.file("extdata", "boundary", "catchments_all_lidar.shp", package = "mrmsR")

list.files(grib_dir)
```

## Step 2: Cropping rasters to a boundary

`prepMRMS()` reads every `.grib2` file in `dir`, assigns the WGS84 CRS,
crops and masks it to `boundary`, and writes the result to `output_dir` as a
GeoTIFF. It skips any file that's already been processed, so it's safe to
rerun on a directory that's still being downloaded into.

```{r prepMRMS}
tif_dir <- file.path(tempdir(), "mrmsR-vignette-tif")
dir.create(tif_dir, showWarnings = FALSE)

prepMRMS(
  dir = grib_dir,
  output_dir = tif_dir,
  num_cores = 1,
  boundary = boundary
)

list.files(tif_dir)
```

```{r plot-raster, fig.alt = "Cropped MRMS precipitation raster over the example catchment boundary"}
r <- terra::rast(list.files(tif_dir, full.names = TRUE)[1])
terra::plot(r, main = names(r))
```

## Step 3: Zonal statistics per catchment

`zonalMRMS()` takes the processed rasters and computes the mean
precipitation within each catchment polygon, writing one CSV per raster to
`output_dir`.

```{r zonalMRMS}
csv_dir <- file.path(tempdir(), "mrmsR-vignette-csv")
dir.create(csv_dir, showWarnings = FALSE)

zonalMRMS(
  raster_dir = tif_dir,
  output_dir = csv_dir,
  boundary = boundary,
  n_workers = 1
)

list.files(csv_dir)
```

```{r zonal-preview}
head(read.csv(list.files(csv_dir, full.names = TRUE)[1]))
```

## Step 4: Combining timesteps into one time series

`combineCSV()` reads every CSV in a directory and combines them into a
single data frame, sorted by `datetime`.

```{r combineCSV}
combined <- combineCSV(csv_dir)

combined
```

From here, `combined` is a normal data frame — grouping by `catchment` and
summarizing over `datetime` with your usual `dplyr`/`tidyr` tools will get
you catchment-level rainfall totals or hyetographs.
